---
title: "Severe Weather Events Analysis Report"
author: Oleg Krivosheev
output:
  html_document:
    keep_md: true
---

# Severe Weather Events Impact on Population Health and Economy

## Synopsis

The United States National Oceanic and Atmospheric Administration's (NOAA) [storm database]
(https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
was used to answer some basic questions about severe weather events
statistics. We identified which types of events are the most harmful to population
health and the most dangerous to the economy. The NOAA's database tracks
characteristics of major storms and weather events in the United States,
including estimates of any fatalities, injuries, and property and crop damage.

Analysing these data, we found out that
tornadoes and heat are the severe weather event types by far most dangerous to people.
Tornados, winds and floods are the most damaging events to the property and crops.
Floods of all kinds are both dangerous to the people and produce
a lot of damage to the property and crops.

## Prerequisites

This report was made using following software:

 - **R** version 3.2.3, 64bit, running under Windows operating system, taken from [R project](https://cran.r-project.org/bin/windows/base/).
 - **data.table** version 1.9.6 from [data table](https://cran.r-project.org/web/packages/data.table/index.html).
 - **ggplot2** as found in the **R** installation.
 - **grid** as found in the **R** installation.
 - **xtable** as loaded from [there](https://cran.r-project.org/web/packages/xtable/index.html).
 - **R Studio** v0.99.489 from [R studio](https://www.rstudio.com/products/rstudio/download/).
 - **7 zip** v15.12, 64bit, from Igor Pavlov [page](http://www.7-zip.org/).

## Data Processing

### Load required **R** packages

We are using **R** packages such as  **data.table**, **xtable** and **ggplot2**. First, we ensure
packages are installed and loaded.

```{r, echo=TRUE}
check_and_install <- function( packname ) {
    # given package name, check if it is installed
    # if not, download and install
    if ( packname %in% rownames(installed.packages()) == FALSE ) {
        install.packages( packname )
    }
}

check_and_install("data.table")
check_and_install("xtable")
check_and_install("ggplot2")
check_and_install("grid")

require("data.table")
require("xtable")
require("ggplot2")
require("grid")
```

### Download and Prepare Data

We download all relevant data, including compressed data table itself together with documentation and FAQ.

```{r, echo=TRUE}

fname.data <- "repdata-data-StormData.csv.bz2"
url.data <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url.data, destfile = fname.data, mode = "wb")

fname.docs <- "repdata-data-StormDocs.pdf"
url.docs <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
download.file(url.docs, destfile = fname.docs, mode = "wb")

fname.faq <- "repdata-data-StormFAQ.pdf"
url.faq <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf"
download.file(url.faq, destfile = fname.faq, mode = "wb")
```

We unpack the NOAA storm database.

```{r, echo=TRUE}
# extractor from Igor Pavlov's 7zip
sevenzip <- "\"C:/Program Files/7-Zip/7z.exe\""
# always overwrite on extract
cmd <- paste(sevenzip, " ", "x", " ", "-aoa", " ", fname.data)
try( system(cmd) )
```

### Load data base

The data file is now in the working directory, and in the well-known CSV format.
Read it directly as **data.table**.

```{r, echo=TRUE}
suppressMessages( sd <- fread("repdata-data-StormData.csv", showProgress = FALSE) )
summary(sd)
```

### Data clean-up

First, we coerce *F* to *integer*.

```{r, echo=TRUE}
# speedy loading
# load("sd.rda")
sd <- sd[, F := as.integer(F)]
```

Second, *BGN_DATE* to be converted to a proper *Date* class type.

```{r, echo=TRUE}
sd <- sd [, BGN_DATE := as.Date(BGN_DATE, format="%m/%d/%Y %H:%M:%S")]
summary(sd)
```

### Data quality versus year investigation

We check how many events are per year total

```{r plot.nof_events_per_year, echo=TRUE}
sd <- sd[, Year := as.POSIXlt(BGN_DATE)$year + 1900]
#str(sd)
sd.per_year <- sd[, .N, by = Year]
#str(sd.per_year)
p <- ggplot(sd.per_year) + geom_line(aes(Year, N)) +
     ylab("Number of events recorded") +
     xlab("Year")
print(p)
```

As one can see, number of stored events is growing as years pass by. That makes analysis
somewhat complicated due to the fact that not all events would be on equal footing.

After deliberation, we decided not to restrict our analysis by year range.

### Economy damage data adjustment

Property damage and crop damage shall be converted into proper numerical form, based upon the
codes described in the code book. Property and crop damages in original data table
are expressed in the different scales using letter codes. Conversion rule is as following.

- If PROPDMGEXP = B, then multiply PROPDMG by 1,000,000,000
- If PROPDMGEXP = M, then multiply PROPDMG by 1,000,000
- If PROPDMGEXP = K, then multiply PROPDMG by 1,000
- If PROPDMGEXP = H, then multiply PROPDMG by 100
- Otherwise leave PROPDMG as-is

Lets check how many NAs are present for property damage data,
and get the number of entries grouped by damage exponent letter code:

```{r, echo=TRUE}
na.PROPDMG <- sum(is.na(sd$PROPDMGEXP))
na.PROPDMGEXP <- sum(is.na(sd$PROPDMGEXP))
```

So far we have `r na.PROPDMG` items with NA for property damage, and
`r na.PROPDMGEXP` items with NA for property damage exponential letter code.

We print the number of appearances of property damage codes in descending order.

```{r, echo=TRUE}
q <- setorder( sd[, .N, by = PROPDMGEXP], -N)
print(q)
```

Same rules applies to the crop damage:

- If CROPDMGEXP = B, then multiply CROPDMG by 1,000,000,000
- If CROPDMGEXP = M, then multiply CROPDMG by 1,000,000
- If CROPDMGEXP = K, then multiply CROPDMG by 1,000
- If CROPDMGEXP = H, then multiply CROPDMG by 100
- Otherwise leave CROPDMG as-is

Check crop damage NA items as well

```{r, echo=TRUE}
na.CROPDMG <- sum(is.na(sd$CROPDMGEXP))
na.CROPDMGEXP <- sum(is.na(sd$CROPDMGEXP))
```

So far we have `r na.CROPDMG` items with NA for crop damage, and
`r na.CROPDMGEXP` items with NA for crop damage exponential letter code.

We print the number of appearances of crop damage codes in descending order.

```{r, echo=TRUE}
print( sum( is.na(sd$CROPDMGEXP) ) )
q <- setorder( sd[, .N, by = CROPDMGEXP], -N)
print(q)
```

We shall define function (below) which will convert damage
exponent letter code into the damage multiplier.

```{r, echo=TRUE}
damage_scale <- function( dmg_code ) {
    #print(dmg_code)
    if ((toupper(dmg_code) == "B")[1]) {
        return( 1.0e9 )
    }
    if ((toupper(dmg_code) == "M")[1]) {
        return( 1.0e6 )
    }
    if ((toupper(dmg_code) == "K")[1]) {
        return( 1000.0 )
    }
    if ((toupper(dmg_code) == "H")[1]) {
        return( 100.0 )
    }

    1.0
}
```

Now, having such function defined, we rescale property damage

```{r, echo=TRUE}
sd <- sd[, PROPDMG := PROPDMG * damage_scale(PROPDMGEXP)]
```

and crop damage as well

```{r, echo=TRUE}
sd <- sd[, CROPDMG := CROPDMG * damage_scale(CROPDMGEXP)]
```

Finally, we add column to our data table which has the value equal to the total damage,
expressed in US dollars.

```{r, echo=TRUE}
sd <- sd[, TOTALDMG := PROPDMG + CROPDMG]
summary(sd)
```

### Population Health data adjustment

We adjust population health related data using very simple consideration.
There are fatalities as well as injuries recorded in the data set. We will
produce common measure of the population health severe events impact by
making data column which combines fatalities weighted by factor of 5 plus
injuries. Such measure will heavily bias fatalities versus injuries, which,
we believe, is a right thing to do in our analysis. We will call it Population Health
Damage index (PHDIDX).

```{r, echo=TRUE}
sd <- sd[, PHDIDX := INJURIES + 5*FATALITIES]
```

### Events aggregation, property damage

We have multiple event types for what is essentially same event. We merge manually several event types
into single with consistent name. We generate such merges for the most significant event types.

First, check number of distinctive event types

```{r, echo=TRUE}
sd.evts <- setorder( sd[, .N, by = .(EVTYPE)], -N )
nof_evttypes <- nrow(sd.evts)
str(sd.evts)
print(sd.evts)
```

We have `r nof_evttypes` distinctive events, but a lot of them are low impact events.

To print most relevant events with regards to the property damage

```{r results="asis", echo=TRUE}
sd.propdmg <- setorder( setnames( sd[ , sum(PROPDMG), by=EVTYPE], "V1", "TOTALDMG" ), -TOTALDMG)
sd.propdmg.sel <- sd.propdmg[1:30]
print(xtable(sd.propdmg.sel, digits = 0), type = "html", include.rownames = FALSE)
```

To check, for example, how many event types are related to hurricanes, search and grep
anything in *EVTYPE* by a part of keyword

```{r, echo=TRUE}
q <- sd.evts[EVTYPE %like% "HURRI"]
print.data.frame(q)
```

We merge all related events into single one by assigning common event type.
For example, we mark as *HURRICANE* all hurricane and typhoon originated events.

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %in% c("HURRICANE/TYPHOON",
                                              "HURRICANE EMILY",
                                              "HURRICANE OPAL",
                                              "HURRICANE ERIN",
                                              "HURRICANE FELIX",
                                              "HURRICANE GORDON",
                                              "HURRICANE OPAL/HIGH WINDS",
                                              "HURRICANE-GENERATED SWELLS",
                                              "TYPHOON"), "HURRICANE", EVTYPE )]
```

Following the established procedure, we combine all thunderstorm winds
related events and mark them as one type of event.

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %in% c(" TSTM WIND",
                                              "TSTM WIND",
                                              "TSTM WIND/HAIL",
                                              "MARINE TSTM WIND",
                                              "SEVERE THUNDERSTORM WINDS",
                                              "THUNDERSTORM WIND",
                                              "THUNDERSTORM WINDS",
                                              "MARINE THUNDERSTORM WIND",
                                              "THUNDERSTORM WINDSS",
                                              "THUNDERSTORM WINDS HAIL",
                                              "THUNDERSTORM WINDS/HAIL",
                                              "THUNDERSTORM WINDS LIGHTNING",
                                              "SEVERE THUNDERSTORM WINDS",
                                              "GUSTY THUNDERSTORM WINDS",
                                              "THUNDERSTORMS WINDS",
                                              "THUNDERSTORM  WINDS",
                                              "THUNDERSTORMS WIND",
                                              "THUNDERSTORM  WINDS.",
                                              "THUNDERSTORMW WINDS",
                                              "THUNDERTORM WINDS",
                                              "THUNDERSTORM WINDS.",
                                              "THUNDERSTROM WINDS",
                                              "THUNDERTSORM WIND",
                                              "THUNDERSTORM WINDS/ HAIL",
                                              "THUNDERSTORMW",
                                              "THUNDERESTORM WINDS",
                                              "THUNDERSTORMWINDS",
                                              "THUNDERSTROM WIND",
                                              "THUNDERSTORM W INDS",
                                              "THUNDERSTORMW 50"
                                               ), "THUNDERSTORM WIND", EVTYPE )]

sd[, EVTYPE := ifelse( toupper(EVTYPE) %like% "THUNDERSTORM WI", "THUNDERSTORM WINDS", EVTYPE )]
```

Aggregate all relevant heavy wind related events into *HEAVY RAIN* event type.

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %in% c("HEAVY RAIN",
                                              "HEAVY RAINS",
                                              "HEAVY RAIN/WIND",
                                              "HEAVY RAINFALL",
                                              "TSTM HEAVY RAIN",
                                              "UNSEASONAL RAIN",
                                              "EXCESSIVE RAINFALL",
                                              "HEAVY RAIN/SEVERE WEATHER"
                                               ), "HEAVY RAIN", EVTYPE )]
```

Aggregate all relevant flash flood related events into *FLASH FLOOD* event type.

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %in% c("FLASH FLOODING",
                                              "FLASH FLOOD/FLOOD",
                                              "FLOOD/FLASH FLOOD"
                                               ), "FLASH FLOOD", EVTYPE )]
```

Similar aggregation wrt the *URBAN FLOOD* event type.

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %in% c("URBAN FLOODING",
                                              "FLOODING",
                                              "URBAN FLOOD",
                                              "URBAN/SML STREAM FLD"
                                               ), "URBAN FLOOD", EVTYPE )]
```

Flooding aggregation

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %in% c("FLOOD",
                                              "FLOODING",
                                               "MAJOR FLOOD"
                                               ), "FLOOD", EVTYPE )]
```

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %in% c("RIVER FLOODING",
                                               "RIVER FLOOD"
                                               ), "RIVER FLOOD", EVTYPE )]
```

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %like% "COASTAL FL",
                       "COASTAL FLOOD", EVTYPE )]

sd[, EVTYPE := ifelse( toupper(EVTYPE) %like% "COASTAL  FL",
                       "COASTAL FLOOD", EVTYPE )]

sd[, EVTYPE := ifelse( toupper(EVTYPE) %like% "CSTL FL",
                       "COASTAL FLOOD", EVTYPE )]
```

Similar aggregation wrt the *HIGH WIND* event type.

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %in% c("HIGH WIND",
                                              "STRONG WIND",
                                              "BLOWING WIND",
                                              "HIGH WINDS",
                                              "STRONG WINDS",
                                              "STRONG WIND"
                                               ), "HIGH WIND", EVTYPE )]
```

Fog aggregation

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %in% c("FOG",
                                              "DENSE FOG"
                                               ), "FOG", EVTYPE )]
```

Hail aggregation

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %in% c("HAIL",
                                              "SMALL HAIL"
                                               ), "HAIL", EVTYPE )]
```

All tropical storms aggregated together

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %like% "TROPICAL STORM",
                       "TROPICAL STORM", EVTYPE )]
```

Snow events aggregation

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %in% c("LIGHT SNOW",
                                              "LAKE-EFFECT SNOW",
                                              "LAKE EFFECT SNOW",
                                              "SNOW",
                                              "MODERATE SNOWFALL",
                                              "RECORD SNOW",
                                              "RECORD SNOWFALL",
                                              "HEAVY SNOW SQUALLS",
                                              "HEAVY LAKE SNOW",
                                              "EXCESSIVE SNOW"
                                               ), "SNOW", EVTYPE )]
```

After aggregation, we print forty most damaging events in the table below.

```{r results="asis", echo=TRUE}
sd.propdmg <- setorder( setnames( sd[ , sum(PROPDMG), by=EVTYPE], "V1", "TOTALDMG" ), -TOTALDMG)
sd.propdmg.sel <- sd.propdmg[1:40]
print(xtable(sd.propdmg.sel, digits = 0), type = "html", include.rownames = FALSE)
```

### Events aggregation, population health damage

```{r results="asis", echo=TRUE}
sd.pubhdmg <- setorder( setnames( sd[ , sum(PHDIDX), by=EVTYPE], "V1", "PHDIDX" ), -PHDIDX)
sd.pubhdmg.sel <- sd.pubhdmg[1:40]
print(xtable(sd.pubhdmg.sel, digits = 0), type = "html", include.rownames = FALSE)
```

We aggregate events related to public health as well

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %in% c("EXCESSIVE HEAT",
                                              "HEAT WAVE",
                                              "RECORD/EXCESSIVE HEAT",
                                              "EXTREME HEAT",
                                              "UNSEASONABLY WARM",
                                              "RECORD HEAT"
                                             ), "HEAT", EVTYPE )]
```

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %in% c("EXTREME COLD",
                                              "FROST/FREEZE",
                                              "FROST",
                                              "DAMAGING FREEZE",
                                              "RECORD COLD",
                                              "COLD/WIND CHILL",
                                              "EXTREME COLD/WIND CHILL",
                                              "UNSEASONABLY COLD",
                                              "HARD FREEZE",
                                              "FREEZE"
                                             ), "COLD", EVTYPE )]
```

```{r, echo=TRUE}
sd[, EVTYPE := ifelse( toupper(EVTYPE) %in% c("WINTER STORMS",
                                              "WINTER STORM"
                                             ), "WINTER STORM", EVTYPE )]
```

## Results

When we analyze the data, we look at total numbers in aggregate.
We don't use an yearly averages due to the fact that
there are many unique events that happen only once
every few years, for example named hurricanes and tropical storms.

### Population Health effects caused by Severe Weather Events

We look at the number of people killed or injured per event type,
adjusted as described in public health index.

For the 40 most dangerous event types:

```{r plot.pop.health.damage, echo=TRUE}
sd.pubhdmg <- setorder( setnames( sd[ , sum(PHDIDX), by=EVTYPE], "V1", "PHDIDX" ), -PHDIDX)
sd.pubhdmg.sel <- sd.pubhdmg[1:40]

plt.phdidx <- ggplot(data=sd.pubhdmg.sel, aes(x = EVTYPE, y = PHDIDX)) +
              theme(plot.margin = unit(c(1,1,-0.2,.91), "cm")) +
              geom_bar(stat="identity") +
              ylab("Population Health Damage Index (5*K+I)")

plt.phdidx <- plt.phdidx + theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
              xlab("")

print(plt.phdidx)
```

And same data are printed in the tabular form

```{r results="asis", echo=TRUE}
print(xtable(sd.pubhdmg.sel[1:30], digits = 0), type = "html", include.rownames = FALSE)
```

We can see from the graph and the table that certain event types are
extremely dangerous to the health of the population, most notably being tornadoes and heat.

### Property and Crop damage caused by Severe Weather Events

We now look at total property and crop damage per event type, for the top most costly event types.

```{r plot.property.crop.damage, echo=TRUE}
sd.propdmg <- setorder( setnames( sd[ , sum(PROPDMG), by=EVTYPE], "V1", "TOTALDMG" ), -TOTALDMG)
sd.propdmg.sel <- sd.propdmg.sel[1:40]

plt.totdmg <- ggplot(data=sd.propdmg.sel, aes(x=EVTYPE, y=TOTALDMG)) +
  theme(plot.margin=unit(c(1,1,-0.2,.82), "cm")) +
  geom_bar(stat="identity") +
  ylab("Property and Crop Damage (USD)")

plt.totdmg <- plt.totdmg +
              theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
              xlab("")

print(plt.totdmg)
```

And data are presented as table as well.

```{r results="asis", echo=TRUE}
print(xtable(sd.propdmg.sel[1:30], digits = 0), type = "html", include.rownames = FALSE)
```

As with dangerous to the human event types,
from the graph and table we can see that certain event
types are extremely costly to the economy.

Heaviest damages are caused by tornados, winds and floods.

## Conclusion

Using **R**, we investigated what kind of severe weather events caused most damage to population
health and the economy. We propose to work with business, goverment and communities to prevent
and alleviate the effects of such events.
