---
title: "Severe Weather Events Analysis Report"
author: Oleg Krivosheev
output:
  html_document:
    keep_md: true
---

Severe Weather Events Impact on population health and Economy,
         Analysis Report
=================

## Synopsis

In this report we aim to answer some basic questions about severe weather events.
Specifically, we try to identify which types of events are the most harmful to population
health and the most deleterious to the economy.
To answer these questions, we obtained the storm database from the U.S. National Oceanic and Atmospheric Administration's (NOAA).
This database tracks characteristics of major storms and weather events in the United States, including estimates of any fatalities,
injuries, and property and crop damage. From these data, we found that tornadoes and heat are the severe weather event types
by far most dangerous to people, while flooding, hurricanes, and storm surges are the most costly event types to the economy.
Interestingly, only flooding is one of the top three most dangerous or most costly event types.

## Prerequisites

This report was produced using following packages

 - **R** version 3.2.3, 64bit, running under Windows operating system, taken from [.](https://cran.r-project.org/bin/windows/base/)
 - **data.table** version 1.9.6 from [data table](https://cran.r-project.org/web/packages/data.table/index.html).
 - **ggplot2** as found in **R** installation
 - **xtable**

## Data Processing


### Loading required **R** packages

```{r, echo=TRUE}
check_and_install <- function( packname ) {
    # given package name, check if it is installed
    # if not, download and install
    if ( packname %in% rownames(installed.packages()) == FALSE ) {
        install.packages( packname )
    }
}

check_and_install("data.table")
check_and_install("xtable")
check_and_install("ggplot2")

require("data.table")
require("xtable")
require("ggplot2")
```

### Reading and Analysing Data

```{r, echo=TRUE}
```




## Assignment

## Results